Supplementary Information:\\Evidence for black-swan events in animal populations

### 2016-01-02 text/ms.tex
Add ms.tex

\textsuperscript{3}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA

\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sean\_anderson@sfu.ca

\textbf{Abstract}:

Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. We derived a model for estimating the frequency of black-swan events and applied it to abundance time series of 609 animal populations. We find strong evidence for black-swan events, but they are rare, occurring in 3--5\% of populations: most frequently for birds (4--8\%), mammals (4--6\%), and insects (2--3\%). Black swan events were predominantly (87\%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases. Black swan events are not explained by any life-history covariates, but tend to be driven by external perturbations such as climate or severe winters. Data and simulations reveal that black swan-events are more likely to be detected in longer and less noisy time series. Our results show that accurate forecasts of extinction risk require accounting for the possibility of black swans and the predominance of crashes over increases, or predicted abundance will underestimate the probability of crashes by a factor of XX--XX. In addition, the presence of black swans highlights the importance of developing flexible and robust conservation and management strategies, particularly as the frequency and magnitude of climate extremes increase over the next century.

Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, may2008}. Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the most damaging earthquakes greatly exceed their expected frequency\cite{sornette2009}. In ecology, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}. Indeed, catastrophes may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}. Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, dennis2006, knape2012}. If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.

Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes. We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes. To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in abundance from one time step to the next are more extreme than typically seen with a lognormal distribution. Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution\emph{.} By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1). For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is \(0.02\), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^-7\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2, Table~S2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a multilevel model, we found stronger evidence for black swans in insect populations than is apparent in Fig.~2---four of eight orders with the highest median probability of heavy tails were insect orders---however, there was considerable uncertainty in these estimates (Fig.~3a).

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (Fig.~S3). Introducing moderate observation error (CV = 0.2) slightly decreased the estimated prevalence of black-swan events (Fig.~S3). The strength of the prior on ν had little influence on estimates of black-swan dynamics (Fig.~S4). Our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy tailed events---especially given the length of the time series in our analysis (Figs S5, S6).

Across populations, the probability of observing black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Figs 3b, 4). Longer time-series length was the strongest covariate of observing black-swan dynamics. For instance, the expected probability density suggesting black-swan dynamics was about 1.6 times greater for a population with 60 time steps compared to one with 30 time steps. However, the absolute change in probability with increased time series length was small (0.20 vs.~0.13 in this example, Fig.~4).

We examined all time series with published explanations of why black-swan events occurred (Table~S2). The majority (87\%) were downward black swans and involved a combination of multiple factors. For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event was a combination of two black-swan events in a row and it took the population three times longer to recover than predicted\cite{stafford1971}. Downward black swans were sometimes followed by upward black swans. For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom, declined suddenly following a red tide event in 1968\cite{potts1980}. This freed quality nest sites for first-time breeders, productivity rapidly increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.

Given the prevalence of downwards events, we refit our heavy-tailed models estimating the skewness of process noise (\(\gamma\); Figs 1a, SX) and used these models to make near-term risk forecasts. Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, SX). In contrast, populations with little evidence of heavy tails (median \(\nu \geq 70\)) also had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these heavy-tailed populations forward five years revealed that assuming lognormal process noise underestimated risk by X--X fold (ratio of interquartile range of 99\% lower credible interval of abundance). \emph{TODO: closing sentence emphasizing the implications of this?}

Thus our analysis provides strong evidence for the presence of downwardly skewed black-swan events in ecological process noise of higher taxa. Unmodelled intrinsic properties of populations or extrinsic forces acting on populations could generate black-swan dynamics. Given that a mixture of normal distributions with different variances can generate a Student-t distribution with heavy tails\cite{allen2001}, we could observe black-swan dynamics if we miss an underlying mixture of processes. Those processes might be an aggregation of populations across space, population diversity\cite{anderson2015}, or an intrinsic change in population variability through time. Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012}. These forces could be extreme themselves. For example, extreme climate, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction might cause black-swan events. Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan ecological dynamics. This could occur with synergistic interactions\cite{kirby2009} or even if non-synergistic forces experience a rare alignment\cite{denny2009}.

There are a number of caveats when considering the generality of our results. The GPDD data represent a taxonomically and geographically biased sample of populations---the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we would expect to find qualitatively similar evidence for black swans in other large taxonomic or geographic samples of populations, the common forces driving those black swans would likely differ. Additionally, some apparent black-swan events could be recording mistakes, or conversely, some extreme observations may have been discarded or altered if they were erroneously suspected of being recording mistakes. Indeed, we discarded three of the populations that our method initially identified as heavy tailed because they turned out to be data-entry errors (Supporting Information). Finally, the temporal scales of observation and population dynamics vary considerably across populations in the GPDD and these likely influence the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large-bodied mammals) we will average across generations and perhaps miss black-swan events. Conversely, if we census populations infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and also dampen black-swan dynamics.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy directions. Ecological resource management can draw from other disciplines that focus on heavy tails. For example, earthquake preparedness and response is focussed on black-swan events. Similarly to ecological black swans, we can rarely predict the specific location and timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}. A similar focus might benefit ecological resource management once our ability to predict the spatial probability and covariates of ecological black swans improves. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances that are appropriately set back from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{hilborn2003, schindler2010, anderson2015}. Finally, extreme and unexpected, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, pine-iii2009, lindenmayer2010}

Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}. In recent decades, ecology has moved toward focussing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}. Financial analysts are concerned with the shape of the downward tails of financial returns because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}. A comparable focus in ecology would increase our estimates of extinction risk, since these would be disproportionately impacted by downward black-swan events. Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.

@@

\bibliographystyle{nature}\bibliography{jshort,ms}

\textbf{Online Content:} Methods, along with any Extended Data display items are available in the online version of the paper; references unique to these sections appear only in the online paper.

\textbf{Acknowledgements}: We thank J.W. Moore, J.D. Yeakel, C. Minto and members of the Earth to Ocean Group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supporting Information. Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).

\textbf{Author Contributions}: S.C.A. and T.A.B. conceived the project; S.C.A., T.A.B., A.B.C., and N.K.D. designed the project; S.C.A. analyzed the data and wrote the paper with input from all authors.

\textbf{Author information}: The authors declare no competing financial interests. Reprints and permissions information is available at . Correspondence and requests for materials should be addressed to S.C.A. (sean\_anderson@sfu.ca).

@@

\textbf{Figure~1: An illustration of fitting population dynamic models that allow for heavy tails, represented by the Student-t degrees of freedom parameter} \(\mathbf{\nu}\)\textbf{. a, b,} The probability density for Student-t distributions with a scale parameter of 1 and different values of \(\nu\). Small values of \(\nu\) create heavy tails. As \(\nu\) approaches infinity the distribution approaches the normal distribution. For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is 0.02, whereas the probability of drawing such a value from a normal distribution is nearly zero (\(2.9 \cdot 10^{- 7}\)). \textbf{c--e,} Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with three different values of \(\nu\). Coloured dots in panels \textbf{c} and \textbf{d} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution. \textbf{f--h,} Estimates of \(\nu\) from models fit to the times series in \textbf{c--e}. Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines). Colour shading behind \textbf{f--h} illustrates the region of heavy tails.

@@

\textbf{Figure~2: Estimates of population dynamic heavy-tailedness for populations of birds, mammals, insects, and fishes.} Small values of \(\nu\) approximately (\(< 10\)) suggest heavy-tailed black-swan dynamics; larger values of \(\nu\) suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that \(\nu < 10\) (probability of heavy tails) for populations arranged by taxonomic order and sorted by decreasing mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in \textbf{a} are omitted for space. Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed black-swan dynamics.

@@

\textbf{Figure~3: Standardized probabilities and covariates of heavy-tailed dynamics.} \textbf{a,} Taxonomic-order-level posterior densities of Pr(\(\nu < 10\)) after accounting for time-series length in a multilevel model. Estimates are at the geometric mean of time series length across all the data (approximately 27 time steps). Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals). Dotted vertical line in \textbf{a} indicates the median expected Pr(\(\nu < 10\)) from the prior distribution. \textbf{b,} Posterior densities for potential covariates of Pr(\(\nu < 10\)). The beta regression models were fit on a logit scale with hierarchical intercepts for taxonomic class, taxonomic order, and species. All covariates were standardized by subtracting their mean and dividing by twice their standard deviation. In both panels, short vertical line segments within the density polygons indicate medians.

@@

\begin{center}
\includegraphics[width=0.55\textwidth]{../analysis/skew-fig.pdf}
\end{center}

\textbf{Figure~4: Evidence for downward skewness in process noise for heavy-tailed populations. a,} Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)). \textbf{b,} Posterior density of the skewness parameters aggregated across all populations grouped into heavy-tailed populations (\(\hat{\nu} < 10\)), slightly heavy-tailed populations (\(10 \leq \hat{\nu} \leq 70\)), and normal tails (\(\hat{\nu} > 70\)). The approximate median posterior density of \(\gamma\) and a mid-value of \(\nu\) from distribution in \textbf{b} are shown in \textbf{a}. \textbf{c,} Example time series of heavy-tailed populations with different levels of skewness. Red dots highlight likely heavy-tailed events. Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance. Black lines indicate forecasts from a standard Gompertz model with lognormal process noise. Red lines indicate forecasts from a Gompertz model with estimated tail shape and skewness.

\textbf{Methods}

We selected abundance time series from the Global Population Dynamics Database (GPDD;\cite{gpdd2010}), which contains nearly 5000 time series of abundance from \(\sim 1000\) species and \(\sim 100\) taxonomic orders. We filtered the data (Supporting materials) to remove populations from less reliable data sources, and those without sufficient data for our models, and then interpolated some missing values (sensu Brook \textit{et al.}\cite{brook2006a}). Our interpolation affected only \(\sim\) \interpPointsPerc \% of the final data points (Supporting materials Table~\ref{tab:stats}) and none of the data points that were later considered black-swan events. Our final dataset contained \NPops\ populations across \NOrders\ taxonomic orders and seven taxonomic classes, with a median of \medianTimeSteps\ time steps (range of \minTimeSteps--\maxTimeSteps) (Extended Data Table~\ref{tab:stats}).

We fit heavy-tailed Gompertz population dynamics models to data from the Global Population Dynamics Database (GPDD). \emph{{[}TODO something about extensive quality control filtering and interpolation of the time series{]}} The Gompertz model represents population growth as a linear function in log space. If \(x_{t}\) represents the log abundance (\(N\)) at time \emph{t} then \(x_t = \lambda + bx_{t - 1} + \epsilon_t,\ \epsilon_t \sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma)\). The growth parameter \(\lambda\) represents the expected growth rate if \(N_{t} = 1\). The model is density independent if \(b = 1\), maximally density dependent if \(b = 0\), and inversely density dependent if \(b < 0\).

Usually, the process noise \(\epsilon_{t}\) is modelled as normally distributed{[}REFs{]}, but we allow it to be drawn from a Student-t distribution with scale parameter \(\sigma\) and degrees of freedom \(\nu\). We can then estimate the degree to which the process deviations have heavy tails and are thereof evidence of black-swan events (Fig.~1a, b).

We chose a weakly informative exponential prior on \(\nu\) with rate parameter of \(0.01\) truncated at values above two---a slightly less informative prior than suggested by Fernandez and Steel\cite{fernandez1998}. This prior gives only a \(0.08\) probability that \(\nu < 10\) but constrains sampling sufficiently to avoid drifting towards infinity. In any case, for \(\nu > 20\) the t distribution is almost indistinguishable from the normal distribution (Fig.~1). Based on the shape of the t distribution, we chose the probability that \(\nu < 10\), Pr(\(\nu < 10\)), to define the probability of heavy-tailed dynamics. When categorizing a population as heavy or normal tailed, we applied a 0.5 probability threshold.

\emph{Details on the heavy-tailed Gompertz probability model}

For the Gompertz model, our weakly-informative priors (Extended Data Fig.~\ref{fig:priors}) were: \(b \sim \mathrm{Uniform}(-1, 2)\), \(\lambda \sim \mathrm{Normal}(0, 10^2)\), \(\sigma \sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\), \(\nu \sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)\). Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We would not expect values of \(b\) with levels of density dependence as low as \(-1\) (very strong inverse density dependence), nor would we generally expect values above \(1\). We allowed values of \(b\) above \(1\) to allow for non-stationary time series of growth rates. The estimates of \(b\) were well within these bounds. Our prior on \(\lambda\) was very weakly informative within the range of expected values for population growth and is similar to the default priors suggested by Gelman \emph{et al.}\cite{gelman2008d} for intercepts of regression models. Our Half-Cauchy prior on \(\sigma\) follows Gelman \emph{et al.}\cite{gelman2006c} and Gelman \emph{et al.}\cite{gelman2008d} and the specific scale parameter of \(2.5\) is based on our expected range of the value in nature from previous studies\cite{connors2014}. In our testing of a subsample of populations, our parameter estimates were not qualitatively changed by switching to an uninformative uniform prior on \(\sigma\), but the weakly informative Half-Cauchy prior sped up chain convergence.

Our prior on \(\nu\) was based on Fernandez and Steel\cite{fernandez1998}. They chose an exponential rate parameter of \(0.1\). We chose a less informative rate parameter of \(0.01\) and truncated the distribution at \(2\), since at \(\nu < 2\) the variance of the t distribution is undefined. This prior gives only a \(7.7\)\% probability that \(\nu < 10\) but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately \(\nu = 20\) the t distribution is so similar to the normal distribution (Fig.~1) that time series of the length considered here are unlikely to be informative about the precise value of \(\nu\). In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~1e, h), the posterior will approximately match the prior (prior median \(= 71\), mean \(= 102\)) and the metrics used in our paper (e.g. \(Pr(\nu < 10) > 0.5\)) are unlikely to flag the population as heavy tailed.

>>>>

\textsuperscript{3}School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA

\textsuperscript{*}Corresponding author: Sean C. Anderson; E-mail: sandrsn@uw.edu; Present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA 98195, USA

%125

\textbf{Abstract}:
Black swans are statistically improbable events that nonetheless occur---often with profound consequences\cite{taleb2007, sornette2009}. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency\cite{meehl2004, katz2005, ipcc2012}, it remains unclear the extent to which ecological populations buffer or suffer from such extremes. Here we derive a model for estimating the frequency and direction of black-swan events and apply it to abundance time series of 609 animal populations. We find strong evidence for black-swan events, but they are rare, occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%). Black-swan events are predominantly (\percBSDown \%) downward, implying that unexpected population crashes occur far more frequently than unexpected population increases. Black-swan events are not explained by any intrinsic life-history covariates, but tend to be driven by extrinsic perturbations such as climate or severe winters. Data and simulations reveal that black-swan events are more likely to be detected in longer and less noisy time series. Rare events tend to be crashes and ignoring them will underestimate extinction risk. Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes. Our results demonstrate the critical importance of allowing for downward and heavy-tailed events in population modelling and developing robust conservation and management strategies, particularly in the face of increased climate extremes.

\bigskip
\bigskip

\textbf{One Sentence Summary}:

\textbf{Main text}:
Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}. Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}. In ecology, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}. Indeed, catastrophes may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}. Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, knape2012}. If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.

Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes (table~S1) \cite{SOM}. We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes. To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in log abundance from one time step to the next are more extreme than typically seen with a normal distribution. Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution. By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1, fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed distributions. For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is moderately high (\(0.02\)), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a hierarchical model, we found stronger evidence for black swans in insect populations than these statistics suggest---four of eight orders with the highest median probability of heavy tails were insect orders (Fig.~3a).

The majority of our heavy-tailed estimates were robust to alternative population models, observation error, and choice of Bayesian priors. Our conclusions were not systematically altered when we included an autocorrelation structure in the residuals, modelled population growth rates with or without density dependence, or modelled the population dynamics as Ricker-logistic instead of Gompertz (fig.~S\ref{fig:alt}). Similarly, introducing moderate observation error (CV = 0.2) only slightly decreased the estimated prevalence of black-swan events (fig.~S\ref{fig:alt}), and the strength of the prior on \(\nu\) had little influence on estimates of black-swan dynamics (fig.~S\ref{fig:alt-priors}). Finally, our simulation testing shows that, if anything, our models underpredict the true magnitude and probability of heavy-tailed events---especially given the time series have a median of only 26 years in our analysis (figs~S\ref{fig:sim-nu},~S\ref{fig:sim-prob}).

For model fits to the population data, the probability of detecting black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}). Longer time-series length was the strongest covariate of observing black-swan dynamics: there is a 1.6 times greater probability of detecting a black swan event in populations with 60 time steps than in one with 30 time steps (fig.~S\ref{fig:correlates}).

We examined all time series with published explanations of why black-swan events occurred (table~S2). The majority (\percBSDown \%) were downward events and involved a combination of multiple factors. For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event was a combination of two heavy-tailed events in a row and the population took three times longer to recover than predicted\cite{stafford1971}. Downward black swans were sometimes followed by upward black swans. For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}. This freed quality nest sites for first-time breeders, productivity increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.

Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness of the process noise using a skew-t distribution (\(\gamma\); Fig.~4a), and used these models to make near-term risk forecasts. Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these heavy-tailed populations forward five years revealed that assuming the standard normal process noise underestimated risk by \crashUnderRange-fold (interquartile range of 99\% lower credible interval of abundance; Fig.~4c, fig.~S\ref{fig:skew-projections}). Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.

There are a number of possible causes of black swan events, including unmodelled intrinsic properties of populations or extrinsic forces acting on populations. For example, we could observe black-swan dynamics if we miss an underlying mixture of processes, since a mixture of normal distributions with different variances can generate a t distribution with heavy tails\cite{allen2001}. In ecology, these processes could arise through an aggregation of populations across space, population diversity\cite{schindler2010}, or an intrinsic change in population variability through time\cite{carpenter2006}. Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012} including extreme climate\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction. Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan dynamics through synergistic interactions\cite{kirby2009} or a rare alignment of non-synergistic forces\cite{denny2009}.

In light of our findings, we suggest that ecological resource management can learn from disciplines that focus on heavy tails. For example, earthquake preparedness and response is focused on black-swan events. Similarly to ecological black swans, we can rarely predict the specific timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances far from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}. Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.

Rare catastrophes can have a profound influence on population persistence\cite{mangel1994}. In recent decades, ecology has moved toward focusing on aspects of variance in addition to mean responses\cite{loreau2010a, thompson2013}. Our results suggest that an added focus on ecological extremes represents the next frontier, particularly in the face of increased climate extremes\cite{meehl2004, ipcc2012, thompson2013}. Financial analysts are concerned with the shape of financial return downward tails because these directly impact estimates of risk---the probability of a specific magnitude of undesired event occurring\cite{rachev2008}. Similarly, ecologists should focus more on estimating and predicting downward tails of population abundance, since these increase true extinction risk.

@@

\bibliographystyle{Science}\bibliography{jshort,ms}

\textbf{Acknowledgements}: We thank J.W. Moore, A.O. Mooers, L.R. Gerber, J.D. Yeakel, C. Minto and members of the Earth to Ocean Group for helpful discussions and comments. We are grateful to the contributors and maintainers of the Global Population Dynamics Database and to Compute Canada's WestGrid high-performance computing resources. Silhouette images were obtained from phylopic.org under Creative Commons licenses; sources are listed in the Supplementary Information. Funding was provided by a Simon Fraser University Graduate Fellowship (S.C.A.), the Natural Sciences and Engineering Research Council of Canada (N.K.D., A.B.C.), the Canada Research Chairs Program (N.K.D.).

\textbf{Supplementary Materials}\\
Materials and Methods\\
References (32 to 72)\\
Tables S1 to S2\\
Figs.~S1 to S8\\

%\textbf{Author Contributions}: S.C.A. and T.A.B. conceived the project; S.C.A., T.A.B., A.B.C., and N.K.D. designed the study; S.C.A. analyzed the data and wrote the paper with input from all authors.

%\textbf{Author information}: The authors declare no competing financial interests. Reprints and permissions information is available at www.nature.com/reprints. Correspondence and requests for materials should be addressed to S.C.A. (sean\_anderson@sfu.ca).

@@

\textbf{Fig.~1: Illustration of population dynamic models that allow for heavy tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t distribution with scale parameter of 1 and different values of \(\nu\). Small values of \(\nu\) create heavy tails. As \(\nu\) approaches infinity the distribution approaches the normal distribution. (\textbf{C} to \textbf{E}) Simulated population dynamics from a Gompertz model with process noise drawn from Student-t distributions with three values of \(\nu\). Coloured dots in panels \textbf{C} and \textbf{D} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution. (\textbf{F} to \textbf{H}) Estimates of \(\nu\) from models fit to the times series in \textbf{C} to \textbf{E}. Shown are posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines). Colour shading behind \textbf{F} to \textbf{H} illustrates the approximate region of heavy tails.

@@

\textbf{Fig.~2: Estimates of population dynamic heavy-tailedness for populations of birds, mammals, insects, and fishes.} (\textbf{A} to \textbf{D}) Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that \(\nu < 10\) for populations arranged by taxonomic order and sorted by decreasing mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in \textbf{A} are omitted for space. Red to yellow points highlight populations with a high to moderately high probability of heavy-tailed dynamics.

@@

\textbf{Fig.~3: Standardized probabilities and covariates of heavy-tailed dynamics.} (\textbf{A}) Taxonomic-order-level posterior densities of Pr(\(\nu < 10\)) after accounting for time-series length in a hierarchical model. Dotted vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the prior distribution. Colour shading refers to taxonomic class (yellow: fishes, green: insects, purple: birds, and red: mammals). Estimates are standardized at the geometric mean of time series length across all the data (approximately 27 time steps) and shown for orders with \(\ge 5\) populations. (\textbf{B}) Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both panels, short vertical line segments within the density polygons indicate medians.

@@

\begin{center}
\includegraphics[width=0.65\textwidth]{../analysis/skew-fig.pdf}
\end{center}

\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)). (\textbf{B}) Posterior density of the skewness parameters aggregated across populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations (\(\hat{\nu} > 70\)). Approximate mid-values from \textbf{B} are illustrated in \textbf{A}. (\textbf{C}) Example time series of heavy-tailed populations with different levels of skewness. Red dots highlight likely heavy-tailed events. Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance. Black and red lines indicate forecasts from Gompertz models with lognormal and skew-t process noise, respectively.
